Supplementary Materials

### 2021-08-27 analysis/paper/paper.Rmd
combine sections to paper.Rmd

---
title: "Untitled"
author:
  - "Bjørn Peare Bartholdy"
  - Amanda G. Henry
date: "`r Sys.Date()`"
output:
  bookdown::pdf_document2:
    df_print: kable
    toc: no
    latex_engine: xelatex
    fig_caption: yes
    number_sections: yes
  bookdown::word_document2: default
  bookdown::html_document2:
    df_print: kable
    keep_md: true
linestretch: 2.0 # double-spacing for review purposes
bibliography: paper/refs.bib
csl: paper/apa.csl
abstract: |
  Lorem ipsum dolor sit amet, consectetur adipiscing elit. Curabitur eget porta erat. Morbi consectetur est vel gravida pretium. Suspendisse ut dui eu ante cursus gravida non sed sem. Nullam sapien tellus, commodo id velit id, eleifend volutpat quam. Phasellus mauris velit, dapibus finibus elementum vel, pulvinar non tellus. Nunc pellentesque pretium diam, quis maximus dolor faucibus id. Nunc convallis sodales ante, ut ullamcorper est egestas vitae. Nam sit amet enim ultrices, ultrices elit pulvinar, volutpat risus.
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.path = "./figures/")
library(tidyverse)
library(broom)
library(here)
library(patchwork)
# load R scripts
source(here("analysis/scripts/analysis.R"))
source(here("analysis/scripts/tables.R"))
source(here("analysis/scripts/figures.R"))
## Compile refs.bib using rbbt package ##
file_paths <- here("analysis/paper", c("intro.Rmd", "matmet.Rmd", "disc.Rmd"))
cite_keys <- sapply(file_paths, rbbt::bbt_detect_citations) 
cite_keys <- unlist(cite_keys, use.names = FALSE)
rbbt::bbt_write_bib("./paper/refs.bib", keys = cite_keys, overwrite = T)
## end refs.bib ##
```

@@

<!-- more detailed background, comprehensible to scientists in related disciplines -->

Until recently, few studies directly investigated the presence of plant microremains
in the dental calculus of archaeological remains. The ability to extract phytoliths
from the dental calculus of archaeological fauna to investigate diet was first 
noted by Armitage [-@armitageExtractionIdentificationOpal1975], 
and later by Middleton and Rovner [-@middletonExtractionOpalPhytoliths1994],
and Fox and colleagues [-@foxPhytolithCalculus1996]. The ability to detect coca
consumption by looking at human dental calculus was explored by Klepinger and Kuhn
[-@klepingerCalculusCoca1977], and starches and phytoliths were extracted from 
human dental calculus by Cummings and Magennis [-@cummingsMayanCalculus1997].

@@

<!-- starches from dental calculus -->
Despite this, our knowledge of dental calculus and the incorporation pathways of 
the various markers is limited [@radiniFoodMultiplePathways2017], as is our 
knowledge of information-loss caused by these pathways and the methods we use to 
extract and analyse dental calculus and make inferences on past diets. 
Extraction methods were tested 
by Tromp and colleagues [-@trompEDTACalculus2017], specifically regarding
decalcification using HCl or EDTA. 
The authors found significantly more starches with the EDTA extraction method 
than the HCl extraction method; however, as noted by the authors, comparisons 
involving archaeological calculus are problematic due to variability between and 
within individuals. Other studies have been conducted on modern populations 
[@leonardDentalCalculus2015] and primates [@powerChimpCalculus2015], both with
well-documented dietary intake, to reconcile our interpretations on the recovery 
of microremains (both phytoliths and starches) from dental calculus and how well
they represent actual dietary intake. These studies are justifiably limited, 
despite meticulous documentation and observation, due to unknown variables and 
uncertainty involved when studying living organisms. Dental calculus is a complex 
oral biofilm with a
multifactorial aetiology and variable formation rates both within and between
individuals [@jepsenCalculusRemoval2011].
Additionally, the effects of the many different methods
of starch processing, as well as post-depositional processes, are still being 
explored [@graneroStarchTaphonomy2020].

<!-- clearly stating the general problem being addressed by this particular study -->
<!-- Dental calculus research has its limitations, all of which I will solve with this
single method. -->
In this exploratory study, we use an oral biofilm model to investigate the 
retention of starch granules within dental calculus, in a controlled laboratory 
setting. Our main questions concern the representation of granules extracted
from the calculus compared to the actual intake. How much of the original diet is 
incorporated into the calculus, and how much is recovered? 
Is there differential loss of information from specific dietary markers that affects 
the obtained dietary information, and how does this affect the representation of diet from
extracted microremains? We find that a limited proportion of the starch input is actually 
retained in the calculus, as well as a shift in the size ratios of individual starch granules
that are incoporated into the calculus. We also show that the number of starch
granules that are incorporated increases as the size of the calculus deposit 
increases.

@@

```{r protocol-fig, fig.cap="Overview of experiment protocol"}
knitr::include_graphics(here("analysis/figures", "ExpL_protocol.png"))
```

<!-- remove sampling from protocol picture -->

@@

Starch treatments were initiated on day 9 to avoid starch granule counts being 
affected by $\alpha$-amylase hydrolysis. An $\alpha$-amylase (EC 3.2.1.1) activity
assay was conducted to confirm that no amylase was present in the system (see 
Bartholdy et al. ???? <!-- ## [bartholdy] ## --> for more details). Starch treatments replaced sucrose treatments, occurring twice per day
for six minutes. The starch treatments involved transferring the substrata to a 
new plate containing a 0.25% (w/v) starch from potato (Roth 9441.1) solution, a 0.25% (w/v) starch from wheat (Sigma S5127) solution, and a 0.5% (w/v) mixture of equal
concentrations (w/v) wheat and potato. All starch solutions were created in a 5%
(w/v) sucrose solution. Before transferring biofilm samples to the starch treatments, 
the starch plates were agitated to keep the starches in suspension in the solutions, 
and during treatments, the rpm was increased to 60.

After 15 days, mineralisation was encouraged with a 
calcium phosphate monofluorophosphate urea (CPMU) solution containing 
20 mmol/l CaCl~2~, 12 mmol/l NaH~2~PO~4~, 5 mmol/l Na~2~PO~3~F, 500 mmol/l Urea
[@pearceConcomitantDepositionStrontium1987; @sissonsMultistationDentalPlaque1991], 
and (0.04 g/l MgCl). The substrata were submerged in 1 ml/well CPMU five times daily, 
every two hours, for six minutes. During the mineralisation period, starch treatments
were reduced to once per day after the five CPMU treatments. This cycle was repeated 
for 10 days until the end of the experiment on day 24 
(see \@ref(fig:protocol-fig) for an overview of protocol). More detailed protocols
are available on protocols.io<!--insert DOI-->.

All laboratory work was conducted in sterile conditions under a laminar flow hood 
to prevent starch and bacterial contamination. Control samples were included to 
detect starch contamination.

@@

<!-- amylase protocol description -->

@@

A 1 ml aliquot of each starch solution was taken, from which 10 $\mu$l was mounted 
on a microscope slide with an 18 x 18 mm coverslip, and counted under a light microscope
(Zeiss Axioscope A1). Counting starches on a full slide was not feasible, so three slide transects were 
counted (at 1/3, 1/2, and 2/3 of the slide), and the sample counts were extrapolated 
(see Supplementary material for more details).

@@

Statistical analysis was conducted in R versions 4.0.5 ("Shake and Throw") 
and 4.0.3 ("Bunny-Wunnies Freak Out") [@R-base].
<!-- R packages used -->
A one-way ANOVA with sample weight as the dependent variable (DV) and treatment 
as the grouping variable (GV) was conducted to explore the effect of the different 
starch treatments on biofilm growth.

@@

Scripts and data will be made available on osf.io upon publication.
Until then, they can be found on the first author's GitHub repository
(https://github.com/bbartholdy/byocStarch).

<!-- script and data availability: provide link (or doi?) to osf.io repository -->

@@

<!-- insert amylase results -->

@@

@@

[@nikitkovaStarchBiofilms2013; @tanStudyBacterialViability2004; @tanCalculusUltrastructure2004].

@@

activity detected in our model, hydrolysis is unlikely to be the main cause of our
reported low incorporation. More details on the absence of $\alpha$-amylase 

@@

<!-- limitations of the platform show which questions would be unanswered -->
This article presents preliminary research exploring the potential of an oral 
biofilm model for dietary research in archaeology. A limitation of this study 
is the absence of amylase in the model. The presence of amylase will likely affect 
the total granule counts as well as the size ratios, as smaller starches may be more 
susceptible to hydrolysis 
[@francoStarchDegradation1992; @haslamDecompositionStarch2004]. 
In future experiments, we recommend the addition of $\alpha$-amylase to the model
in concentrations similar to those found in the oral cavity <!-- provide concentration and citation.--> 
<!-- foreshadowing... -->
While we are able to show the differing 
incorporation caused by absolute counts available, this
was merely a side-effect of the difference in the number of granules in potato and 
wheat solutions of the same concentration (w/v). Further research should test 
multiple differing concentrations of the same starch type.
The use of EDTA may also have affected counts. While previous studies have shown 
negligible morphological changes caused by exposure to EDTA 
[@trompEDTACalculus2017; @lemoyneCalculusPretreatments2021; @modiCalculusMethodologies2020], 
these studies have not considered changes to separate size
categories within starch types, and whether shifts in size ratios occur due to
exposure to the pre-treatment chemicals.
The total number of granules on a slide often exceeded a number that 
was feasible to count in a reasonable time period, so we calculated the total 
counts by extrapolating from three slide transects. 
Thus, we reasonably assume that the three transects are a good representation 
of the entire slide, and that the distribution of all granules on the slide is 
relatively homogenous.   
Finally, we only used native starches in the experimental procedure and the results 
will likely differ for processed starches [@graneroStarchTaphonomy2020]. 
Based on counts obtained by Leonard
and colleagues [-@leonardDentalCalculus2015, Supplement 2], processing and amylase
may have a substantial effect on starch granule retention in the oral cavity.

<!-- How this engineering will facilitate future research -->
The oral biofilm model described in this study, and in 
Bartholdy and colleagues (in prep.),<!-- insert citation --> provides a method to directly
explore the incorporation and extraction of dietary compounds from dental calculus 
in a controlled laboratory setting. The addition of known species in separate
samples also removes the identification bias against the smaller undiagnostic starches.<!-- rephrase -->
It can also address the call for more baseline testing of biases associated with 
dietary research conducted on dental calculus 
[@lemoyneCalculusPretreatments2021].
<!-- How it will facilitate the specific topic of research (e.g. archaeology) -->
*In vivo* studies on humans and non-human primates will be limited by the variability 
occurring
between individuals, and difficulties documenting the entirety of starch granules
entering the oral cavity at any given moment, especially accidental intake. 
Our experimental setup allows us a
higher degree of control over the factors that influence starch incorporation, such as
dietary intake, processing of starches leading to differential survivability, 
and inter- and intra-individual variation in plaque accumulation and 
mineralisation.
The latter is especially difficult to control *in vivo* as it is influenced by 
numerous factors including genetics, diet, and tooth morphology [@jepsenCalculusRemoval2011]<!-- more citations -->.  
Further research is necessary to explore the mechanisms of the intra-oral starch 
incorporation and retention in dental calculus, and the dental calculus model
presented in this study is uniquely suited to explore these questions and 
may improve interpretations of dietary research in past populations.
<!-- still need to replicate results and do more science, but pretty cool, right? -->

@@

Our preliminary study shows that a very small proportion of the ingested starch
granules are retained in a dental calculus model.

Calculus has a low capacity for retention of starch granules, and this effect is
compounded by diagenetic effects in archaeological remains, resulting in low overall
counts of extracted granules. Additionally, the size of granules will influence the 
likelihood of incorporation, as large starches have a decreased incorporation, medium
starches increased, and small starches the only representative of intake.

Given the direct proportional incorporation of starches with quantity of exposure<!-- rephrase -->, 
we can say that the proportion of starches consumed will in many cases be reflected 
in the quantity
of starches extracted from the dental calculus, at least in modern calculus samples
unaffected by diagenesis.
Whether or not this also applies to archaeological samples remains to be tested.

The size of calculus deposit also influences the maximum capacity of granule incorporation; as the size of the deposit increases, so does the absolute count of
incorporated granules.

The data from this study are not sufficient to confirm or deny any proposed methods
of starch incorporation; however, while we are unable to address the mechanism(s) of
starch incorporation with the current experimental setup, this is something that can
potentially be addressed with the calculus model presented in this paper.

<!-- Forward-looking statements about the field as a whole -->
While we have shown multiple factors that influence the likelihood 
of incorporation, it seems the process is somewhat stochastic, and possibly related 
to the biofilm mineralisation events (as discussed above), which are irregular and
related to various factors<!-- citation needed -->. 
This means that the dietary components extracted from dental calculus provide a 
snapshot of the diet around the time of the mineralisation event, which may not 
be an accurate picture of the diet as a whole.

The dental calculus biofilm model presented in this study is a powerful tool to 
investigate various aspects
of dietary research.

>>>>

---
title: "Untitled"
author:
  - "Bjørn Peare Bartholdy"
  - Amanda G. Henry
date: "`r Sys.Date()`"
output:
  bookdown::pdf_document2:
    #template: "../templates/sn-article-template/sn-article.tex"
    df_print: kable
    toc: no
    latex_engine: xelatex
    fig_caption: yes
    number_sections: yes
  bookdown::html_document2:
    df_print: kable
    keep_md: true
  bookdown::word_document2:
    reference_docx: "../templates/template.docx"
linestretch: 2.0 # double-spacing for review purposes
bibliography: refs.bib
csl: apa.csl
abstract: |
  Lorem ipsum dolor sit amet, consectetur adipiscing elit. Curabitur eget porta erat. Morbi consectetur est vel gravida pretium. Suspendisse ut dui eu ante cursus gravida non sed sem. Nullam sapien tellus, commodo id velit id, eleifend volutpat quam. Phasellus mauris velit, dapibus finibus elementum vel, pulvinar non tellus. Nunc pellentesque pretium diam, quis maximus dolor faucibus id. Nunc convallis sodales ante, ut ullamcorper est egestas vitae. Nam sit amet enim ultrices, ultrices elit pulvinar, volutpat risus.
keywords: |
  keyword 1; keyword 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.path = "../figures/")
library(tidyverse)
library(broom)
library(here)
library(patchwork)
# load R scripts
source(here("analysis/scripts/analysis.R"))
source(here("analysis/scripts/tables.R"))
source(here("analysis/scripts/figures.R"))
source(here("analysis/scripts/amylase-assay.R"))
## Compile refs.bib using rbbt package ##
cite_keys <- rbbt::bbt_detect_citations(here("analysis/paper/paper.Rmd")) 
#cite_keys <- unlist(cite_keys, use.names = FALSE)
rbbt::bbt_write_bib(here("analysis/paper/refs.bib"), keys = cite_keys, overwrite = T)
## end refs.bib ##
```

@@

<!-- more detailed background, comprehensible to scientists in related disciplines -->
Until recently, few studies directly investigated the presence of plant microremains
in the dental calculus of archaeological remains. The ability to extract phytoliths
from the dental calculus of archaeological fauna to investigate diet was first 
noted by Armitage [-@armitageExtractionIdentificationOpal1975], 
and later by Middleton and Rovner [-@middletonExtractionOpalPhytoliths1994],
and Fox and colleagues [-@foxPhytolithCalculus1996]. The ability to detect coca
consumption by looking at human dental calculus was explored by Klepinger and Kuhn
[-@klepingerCalculusCoca1977], and starches and phytoliths were extracted from 
human dental calculus by Cummings and Magennis [-@cummingsMayanCalculus1997].

@@

<!-- clearly stating the general problem being addressed by this particular study -->
<!-- Dental calculus research has its limitations, all of which I will solve with this
single method. -->
Despite this, our knowledge of dental calculus and the incorporation pathways of 
the various markers is limited [@radiniFoodMultiplePathways2017], as is our 
knowledge of information-loss caused by these pathways and the methods we use to 
extract and analyse dental calculus and make inferences on past diets. 
Extraction methods were tested 
by Tromp and colleagues [-@trompEDTACalculus2017], specifically regarding
decalcification using HCl or EDTA. 
The authors found significantly more starches with the EDTA extraction method 
than the HCl extraction method; however, as noted by the authors, comparisons 
involving archaeological calculus are problematic due to variability between and 
within individuals. Other studies have been conducted on modern populations 
[@leonardDentalCalculus2015] and primates [@powerChimpCalculus2015], both with
well-documented dietary intake, to reconcile our interpretations on the recovery 
of microremains (both phytoliths and starches) from dental calculus and how well
they represent actual dietary intake. These studies are justifiably limited, 
despite meticulous documentation and observation, due to unknown variables and 
uncertainty involved when studying living organisms. Dental calculus is a complex 
oral biofilm with a multifactorial aetiology and variable formation rates both 
within and between individuals [@jepsenCalculusRemoval2011], contributing to
the stochasticity of starch representation being observed in numerous studies.
Additionally, the rate of $\alpha$-amylase differs both between and within
individuals [@froehlichEffectOral1987; @naterHumanAmylase2005],
causing different rates of hydrolysis of the starch granules present in the oral
cavity. Add to this the effects of the many different methods
of starch processing, as well as post-depositional processes that are still being 
explored [@graneroStarchTaphonomy2020], and you have a highly unpredictable process.

In this exploratory study, we use an oral biofilm model to investigate the 
retention of starch granules within dental calculus in a controlled laboratory 
setting, allowing us full control over dietary input. We use a multiwell setup for consistent, high-throughout analysis.
Our main questions concern the representation of granules extracted
from the calculus compared to the actual intake. How much of the original diet is 
incorporated into the calculus, and how much is recovered? 
Is there differential loss of information from specific dietary markers that affects 
the obtained dietary information, and how does this affect the representation of 
diet from extracted microremains?  
We find that, despite an absence of $\alpha$-amylase in
the model, a limited proportion of the starch input is actually 
retained in the calculus, as well as a shift in the size ratios of individual 
starch granules
that are incoporated into the calculus. We also show that the number of starch
granules that are incorporated increases as the size of the calculus deposit 
increases.

@@

```{r protocol-fig, fig.cap="Overview of experiment protocol"}
knitr::include_graphics(here("analysis/figures", "ExpL_protocol.png"))
```
<!-- modify sampling days from protocol picture to match amylase sampling -->

@@

Starch treatments were initiated on day 9 to avoid starch granule counts being 
affected by $\alpha$-amylase hydrolysis. An $\alpha$-amylase (EC 3.2.1.1) activity
assay was conducted to confirm that no amylase was present in the system following
starch treatments. Starch treatments replaced sucrose treatments, occurring twice per day
for six minutes. The starch treatments involved transferring the substrata to a 
new plate containing a 0.25% (w/v) starch from potato (Roth 9441.1) solution, a 0.25% (w/v) starch from wheat (Sigma S5127) solution, and a 0.5% (w/v) mixture of equal
concentrations (w/v) wheat and potato. All starch solutions were created in a 5%
(w/v) sucrose solution. Before transferring biofilm samples to the starch treatments, 
the starch plates were agitated to keep the starches in suspension in the solutions, 
and during treatments, the rpm was increased to 60.

After 15 days, mineralisation was encouraged with a 
calcium phosphate monofluorophosphate urea (CPMU) solution containing 
20 mmol/l CaCl~2~, 12 mmol/l NaH~2~PO~4~, 5 mmol/l Na~2~PO~3~F, 500 mmol/l Urea, 
and (0.04 g/l MgCl)
[@pearceConcomitantDepositionStrontium1987; @sissonsMultistationDentalPlaque1991].
The substrata were submerged in 1 ml/well CPMU five times daily, 
every two hours, for six minutes. During the mineralisation period, starch treatments
were reduced to once per day after the five CPMU treatments. This cycle was repeated 
for 10 days until the end of the experiment on day 24 
(see \@ref(fig:protocol-fig) for an overview of the protocol). More detailed protocols
are available on protocols.io<!--insert DOI-->.

All laboratory work was conducted in sterile conditions under a laminar flow hood 
to prevent starch and bacterial contamination. Control samples, that only received
sucrose as a treatment, were included to detect starch contamination.

@@

An $\alpha$-amylase (EC 3.2.1.1) activity assay was conducted on artificial
saliva samples collected from the plate wells on days 3, 5, 6, 9, and 12. 
Whole saliva samples were collected on days 0, 3, and 5 as positive controls.
Collected samples were stored at 4&deg;C until the assay was conducted on day 18. 
All samples and standard curves were added in triplicates to two separate plates.
Positive control saliva samples were compared against a standard curve containing 
H~2~O, while artificial saliva samples were compared against a standard curve
containing sterile artificial saliva (due to the colour of artificial saliva).
Two photometric readings were conducted for each plate with a 540 nm filter on a... 
<!-- photometric reader specs -->
The protocol is a slightly modified version of an Enzymatic Assay of $\alpha$-Amylase
(https://www.sigmaaldrich.com/NL/en/technical-documents/protocol/protein-biology/enzyme-activity-assays/enzymatic-assay-of-a-amylase) [@bernfeldAmylase1955], which measures the amount of 
maltose released from starch by $\alpha$-amylase activity.
For a more detailed protocol, see <!-- protocols.io DOI -->

@@

A 1 ml aliquot of each starch solution was taken, from which 10 $\mu$l was mounted 
on a microscope slide with an 18 x 18 mm coverslip, and counted under a light microscope
(Zeiss Axioscope A1). Counting starches on a full slide was not feasible, so three slide transects were 
counted (at ca. 1/4, 1/2, and 3/4 of the slide), and the sample counts were extrapolated 
(see Supplementary material for more details).

@@

Statistical analysis was conducted in `r version$version.string`  [@R-base].
<!-- R packages used -->
A one-way ANOVA with sample weight as the dependent variable (DV) and treatment 
as the grouping variable (GV) was conducted to explore the effect of the different 
starch treatments on biofilm growth.

@@

All scripts and data are available on OSF (https://osf.io/uc5qy/)
and Github (https://github.com/bbartholdy/byoc-starch).

@@

```{r amylase-activity}
# table of results reported in units amylase per mL enzyme (but let's be honest,
  # ...it doesn't really matter what the unit is. No activity is no activity)
cols <- c("1", "2", "3")  
rows <- c("S1", "S2", "S3", "B1", "B2", "B3", "B4", "B5", "BT1", "BT2", "BT3")
plt1_ph1_result <- rbind(sal1_ph1, bmm1_ph1)
  plt1_ph1_result[plt1_ph1_result < 0] <- 0
```

```{r amylase-activity-tbl}
rownames(plt1_ph1_result) <- rows
colnames(plt1_ph1_result) <- cols
knitr::kable(signif(plt1_ph1_result, 3), row.names = TRUE, caption = "Amylase activity in U/mL enzyme, where a U is mg maltose released from starch in six minutes at 36 degrees Celsius. Plate 1, photometric reading 1. Negative values converted to 0.")
```

No appreciable amount of $\alpha$-amylase activity was detected in any of the artificial
saliva samples from any of the days that were sampled. Only positive controls
contained amylase activity that could be detected in the assay (Table
\@ref(tab:amylase-activity-tbl)). The results from the additional plates and
photometric readings can be found in the 
[Supplementary Materials](../supplementary-materials/supp-mat.Rmd). The results
are not comparable to other studies presenting $\alpha$-amylase activity levels
in humans; however, they are sufficient to show that there is no activity in the
system.

@@

<!-- INSERT PICTURES OF STARCHES -->

@@

[@scannapiecoSalivaryAmylase1993; @nikitkovaStarchBiofilms2013; @tanStudyBacterialViability2004; @tanCalculusUltrastructure2004].

@@

activity detected in our model, hydrolysis is not the cause of the low
incorporation rate. More details on the absence of $\alpha$-amylase 

@@

<!-- limitations of the platform show which questions would be unanswered -->
This article presents preliminary research exploring the potential of an oral 
biofilm model for dietary research in archaeology. A limitation of this study 
is the absence of amylase in the model. The presence of amylase will likely affect 
the total granule counts as well as the size ratios, as smaller starches may be more 
susceptible to hydrolysis 
[@francoStarchDegradation1992; @haslamDecompositionStarch2004]. 
In future experiments, we recommend the addition of $\alpha$-amylase to the model
in concentrations similar to those found in the oral cavity [@scannapiecoSalivaryAmylase1993]. 
<!-- foreshadowing...? -->
While we are able to show the differing 
incorporation caused by absolute counts available, this
was merely a side-effect of the difference in the number of granules in potato and 
wheat solutions of the same concentration (w/v). Further research should test 
multiple differing concentrations of the same starch type.
The use of EDTA may also have affected counts. While previous studies have shown 
negligible morphological changes caused by exposure to EDTA 
[@trompEDTACalculus2017; @lemoyneCalculusPretreatments2021; @modiCalculusMethodologies2020], 
these studies have not considered changes to separate size
categories within starch types, and whether shifts in size ratios occur due to
exposure to the pre-treatment chemicals.
The total number of granules on a slide often exceeded a number that 
was feasible to count in a reasonable time period, so we calculated the total 
counts by extrapolating from three slide transects. 
Thus, we reasonably assume that the three transects are a good representation 
of the entire slide, and that the distribution of all granules on the slide is 
relatively homogenous.   
Finally, we only used native starches in the experimental procedure and the results 
will likely differ for processed starches [@graneroStarchTaphonomy2020]. 
Based on counts obtained by Leonard
and colleagues [-@leonardDentalCalculus2015, Supplement 2], processing and amylase
may have a substantial effect on starch granule retention in the oral cavity.

<!-- How this engineering will facilitate future research -->
The oral biofilm model described in this study, and in 
Bartholdy and colleagues (in prep.),<!-- insert citation --> provides a method
to directly
explore the incorporation and extraction of dietary compounds from dental calculus 
in a controlled laboratory setting. The addition of known starch species in
separate samples also removes the identification bias against the smaller
undiagnostic starches.<!-- rephrase -->
It can also address the call for more baseline testing of biases associated with 
dietary research conducted on dental calculus 
[@lemoyneCalculusPretreatments2021].
<!-- How it will facilitate the specific topic of research (e.g. archaeology) -->
*In vivo* studies on humans and non-human primates will be limited by the variability 
occurring between individuals, and difficulties in accurately documenting 
the entirety of starch granules entering the oral cavity at any given moment,
especially accidental intake. 
Our experimental setup allows us a
higher degree of control over the factors that influence starch incorporation, such as
dietary intake, processing of starches leading to differential survivability, 
and inter- and intra-individual variation in plaque accumulation and 
mineralisation.
The latter is especially difficult to control *in vivo* as it is influenced by 
numerous factors including genetics, diet, and tooth morphology [@jepsenCalculusRemoval2011]<!-- more citations -->.  
Further research is necessary to explore the mechanisms of the intra-oral starch 
incorporation and retention in dental calculus, and the dental calculus model
presented in this study is uniquely suited to explore these questions and 
may improve interpretations of dietary research in past populations.
<!-- still need to replicate results and do more science, but pretty cool, right? -->

@@

This preliminary study shows that a very small proportion of the input starch
granules are retained in a dental calculus model. This and previous studies 
have shown that calculus has a low capacity for retention of starch granules, 
an effect that is compounded by diagenetic effects in archaeological remains, 
resulting in low overall counts of extracted granules. 
The proportion of starches consumed will in many cases be reflected 
in the quantity
of starches extracted from the dental calculus---i.e. the more starch granules
entering the oral cavity, the more will be recovered from extraction---at 
least in modern calculus samples unaffected by diagenesis and hydrolysis.
Whether or not this also applies to archaeological samples remains to be tested.
Additionally, we have
shown that the size of granules will influence the likelihood of incorporation, 
as large starches have a decreased incorporation rate, medium
starches an increased rate, and small starches remained somewhat constant.
The size of calculus deposit also seems to influence the capacity of 
granule incorporation; as the size of the deposit increases, so does the
absolute count of incorporated granules.
<!--The data from this study are not sufficient to confirm or deny any proposed
methods of starch incorporation; however, while we are unable to address the 
mechanism(s) of starch incorporation with the data obtained in this study, 
this is something that can potentially be addressed with the calculus model 
presented in this paper.-->

<!-- Forward-looking statements about the field as a whole -->
While we have shown multiple factors that influence the likelihood 
of incorporation, it seems the process is somewhat stochastic, and possibly related 
to the biofilm mineralisation events (as discussed above), which are irregular and
related to various factors. 
The dental calculus biofilm model presented in this study is a powerful tool to 
investigate various aspects of dietary research and starch incorporation and
retention in dental calculus.

### 2021-08-27 analysis/supplementary-materials/supp-mat.Rmd
combine sections to paper.Rmd

>>>>

## Amylase activity

Tables containing the amylase activity results for both plates and photometric
readings.

```{r amylase-activity-results}
# table of results reported in units amylase per mL enzyme (but let's be honest,
  # ...it doesn't really matter what the unit is. No activity is no activity)
cols <- c("1", "2", "3")  
rows <- c("S1", "S2", "S3", "B1", "B2", "B3", "B4", "B5", "BT1", "BT2", "BT3")
(plt1_ph1_result <- rbind(sal1_ph1, bmm1_ph1))
#  plt1_ph1_result[plt1_ph1_result < 0] <- 0
(plt1_ph2_result <- rbind(sal1_ph2, bmm1_ph2))
#  plt1_ph2_result[plt1_ph2_result < 0] <- 0
(plt2_ph1_result <- rbind(sal2_ph1, bmm2_ph1))
  #plt2_ph1_result[plt2_ph1_result < 0] <- 0
(plt2_ph2_result <- rbind(sal2_ph2, bmm2_ph2))
  #plt2_ph2_result[plt2_ph2_result < 0] <- 0
```
